A forest is more than its trees: haplotypes and inferred ARGs

### 2024-10-21 paper/writeup.tex
new figures p2

% Description of algorithm
Given a tree sequence, our goal is to
identify areas of implied inheritance of haplotypes.
Generalizing from \figref{fig:conceptual} (or \figref{fig:extending_diagram}?),
we do this by identifying paths of inheritance that are shared across a sequence of marginal trees
but for which some of the intermediate nodes are missing.
Concretely, suppose that 
if in tree $T_k$ there is a chain of inheritance
$p \to u_1 \to \cdots \to u_m \to c$
(where $a \to b$ denotes a parent-child relationship)
and in tree $T_{k+1}$ there is a chain of inheritance
$p \to v_1 \to \cdots \to v_n \to c$,
where $\{u_i\}_{i=1}^m$ and $\{v_j\}_{j=1}^n$ are disjoint.
This situation implies that $c$ inherited from $p$ over the entire interval $[a_{k-1}, a_{k+1})$,
so it seems reasonable to assume that $c$ has inherited from $p$ \emph{along the same path} for that entire interval.
In other words, the intermediate nodes $\{u_i\}$ should also lie on the path from $c$ to $p$ in tree $T_{k+1}$,
and conversely the nodes $\{v_j\}$ should lie on that path in tree $T_k$.
Of course, this does not always make sense --
for instance, if $u_i$ is already represented somewhere else in $T_{k+1}$,
or if $t_{u_i} = t_{v_j}$ for some $i$ and $j$.
for which the intermediate nodes are not in the other tree
So, we restrict our attention to pairs of such paths in adjacent trees
for which
$u_i \notin T_{k+1}$ for all $1 \le i \le m$,
$v_j \notin T_k$ for all $1 \le j \le n$,
and the times of the nodes $\{u_i\}$ and $\{v_j\}$ are unique.
Call a pair of such paths \emph{mergeable}.
So, the goal of our algorithm is to iterate over trees,
identify mergeable pairs of paths,
and then extend the nodes $\{u_i\}$ to $T_{k+1}$.
(We also extend $\{v_j\}$ to $T_k$, but on a backwards pass.)

\begin{figure}[!ht]
\begin{center}
	\includegraphics[width=5in]{edge_extend_method.pdf}
\end{center}
\caption{\comment{Change figure to say Extend halplotypes} A visualization of the \textit{extend haplotypes} method. Given two neighboring trees $T_1$ and $T_2$, $F_O$ in the tree $T_1$ contains the edge $3\to 0$. Starting at $0$, there is a path of edges $0 \to 3 \to 4$. The tree $T_2$ has edge $4 \to 0$ in $F_I$, and so there exists path $0 to 4$. These two paths share common end points, $0$ and $4$, and so we will extend path $0 \to 3 \to 4$ onto tree $T_2$, hence removing the edge $4\to 0$.
    \label{fig:extending_diagram}
}
\end{figure}

\begin{figure}[!ht]
	\begin{center}
		\includegraphics[width=0.4\linewidth, height=1.2in]{EH-Ex2.pdf}
		\includegraphics[width=0.4\linewidth, height=1.2in]{EH-extended-Ex2.pdf}
	\end{center}
	\caption{\comment{Change figure to have the same arrow with 'Extend Haplotypes' in inkscape for unity's sake.}The extend haplotypes method between tree with more unary nodes. Given two neighboring trees, there exists paths $4\to 6\to8\to10$ and $5\to 9\to 10$ in $F_O$, and in $F_I$ we have paths $4\to 7\to 11\to 10$ and $5\to 12\to 10$. Both the out-forest and in-forest both contain paths with the same initial and terminal nodes, $4\to 10$ and $5\to 10$ respectively, we merge both paths and output the tree sequence on the right.}
	\label{fig:complicated_extending_diagram}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_notlog.pdf}
	\caption{Correct Unary Span for Extend Haplotypes}
	\label{fig:correct-unary-span}
\end{figure}

>>>>

% Description of algorithm
Given a tree sequence, our goal is to
identify areas of implied inheritance of haplotypes.
Generalizing from \figref{fig:extending_diagram},
we do this by identifying paths of inheritance that are shared across a sequence of marginal trees
but for which some of the intermediate nodes are missing.
Concretely, suppose that 
if in tree $T_k$ there is a chain of inheritance
$p \to u_1 \to \cdots \to u_m \to c$
(where $a \to b$ denotes a parent-child relationship)
and in tree $T_{k+1}$ there is a chain of inheritance
$p \to v_1 \to \cdots \to v_n \to c$,
where $\{u_i\}_{i=1}^m$ and $\{v_j\}_{j=1}^n$ are disjoint.
This situation implies that $c$ inherited from $p$ over the entire interval $[a_{k-1}, a_{k+1})$,
so it seems reasonable to assume that $c$ has inherited from $p$ \emph{along the same path} for that entire interval.
In other words, the intermediate nodes $\{u_i\}$ should also lie on the path from $c$ to $p$ in tree $T_{k+1}$,
and conversely the nodes $\{v_j\}$ should lie on that path in tree $T_k$.
Of course, this does not always make sense --
for instance, if $u_i$ is already represented somewhere else in $T_{k+1}$,
or if $t_{u_i} = t_{v_j}$ for some $i$ and $j$.
for which the intermediate nodes are not in the other tree
So, we restrict our attention to pairs of such paths in adjacent trees
for which
$u_i \notin T_{k+1}$ for all $1 \le i \le m$,
$v_j \notin T_k$ for all $1 \le j \le n$,
and the times of the nodes $\{u_i\}$ and $\{v_j\}$ are unique.
Call a pair of such paths \emph{mergeable}.
So, the goal of our algorithm is to iterate over trees,
identify mergeable pairs of paths,
and then extend the nodes $\{u_i\}$ to $T_{k+1}$.
(We also extend $\{v_j\}$ to $T_k$, but on a backwards pass.)

\begin{figure}[!ht]
\begin{center}
	\includegraphics[width=5in]{edge_extend_method.pdf}
\end{center}
\caption{A visualization of the \textit{extend haplotypes} method. Given two neighboring trees $T_1$ and $T_2$, there exists paths from root node $6$ to sample node $0$. $T_1$ contains the path $6\to 4\to 0$ while $T_2$ has path $6\to 5\to 0$. The intermediate nodes $4$ and $5$ do not appear in $T_2$ respectively $T_1$, and so we consider these two paths \textit{mergeable}. The extend haplotypes method joins these two paths, constructing the merged path $6\to 5\to 4\to 0$ for both $T_1$ and $T_2$.
    \label{fig:extending_diagram}
}
\end{figure}

\begin{figure}[!ht]
	\begin{center}
		\includegraphics[width=0.4\linewidth, height=1.2in]{EH-Ex2.pdf}
		\includegraphics[width=0.4\linewidth, height=1.2in]{EH-extended-Ex2.pdf}
	\end{center}
	\caption{\comment{DELETE}}
%	\caption{\comment{Change figure to have the same arrow with 'Extend Haplotypes' in inkscape for unity's sake.}The extend haplotypes method between tree with more unary nodes. Given two neighboring trees, there exists paths $4\to 6\to8\to10$ and $5\to 9\to 10$ in $F_O$, and in $F_I$ we have paths $4\to 7\to 11\to 10$ and $5\to 12\to 10$. Both the out-forest and in-forest both contain paths with the same initial and terminal nodes, $4\to 10$ and $5\to 10$ respectively, we merge both paths and output the tree sequence on the right.}
	\label{fig:complicated_extending_diagram}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_notlog.pdf}
	\caption{\comment{DELETE}}
%	\caption{Correct Unary Span for Extend Haplotypes}
	\label{fig:correct-unary-span}
\end{figure}
